Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-10-01 submission/NatCom_methods.tex
pass to address most of the presubmission comments for nat comm

Consider the objective/optimization function (parameterized by $\mathbf{W}_{l}$s and $\mathbf{b}_{l}$s) for a DNN with $L$ layers, activation functions $h_{l}(\cdot)$, and $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$, as 
the minimization of a general Loss Function $\mathcal{L}$ over $N$ training data instances and labels, $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$.
For a typical supervised classification problem, one has
\begin{equation}
\underset{\mathbf{W}_{l},\mathbf{b}_{L}}{\text{argmin}}\;\sum_{i=1}^{N}\left(\mathcal{L}(E_{DNN}(\mathbf{x}_{i})-y_{i})\right)  ,
\end{equation}
where the loss function $\mathcal{L}(\cdot)$ can take on a myriad of forms~\cite{JC17_TR}, and where
\begin{equation}
E_{DNN}=h_{L}(\mathbf{W}_{L}\times h_{L-1}(\mathbf{W}_{L-1}\times h_{L-2}(\cdots)+\mathbf{b}_{L-1})+\mathbf{b}_{L})
\label{eqn:dnn_energy}
\end{equation}
is the Energy (or optimization) Landscape function.  
$E_{DNN}$ does not explicitly depend on the data, but rather it maps data instance vectors ($\mathbf{x}_i$ values) to predictions (labels).
Therefore, one can analyze $E_{DNN}$ in the absence of any training or test data.

Test accuracies have been reported online for publicly-available pretrained pytroch models.
These models have been trained and evaluated on labeled data $\{\mathbf{x}_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
We do not have access to this data, and we have not trained any of the models ourselves. 
Our methodological approach is thus similar to a statistical meta-analysis, common in biomedical research, but uncommon in ML.
%
Computations were performed with the publicly-available \emph{WeightWatcher} tool (version 0.2.7)~\cite{weightwatcher_package}.
To be fully reproducible, we only examine publicly-available, pretrained models, and we provide all Jupyter and Google Colab notebooks used in an accompanying github repository~\cite{kdd20_sub_repo}.
See Appendix~\ref{sxn:appendix} for details.

\paragraph{Metrics for DNN Weight Matrices.}

Our approach involves analyzing DNN weight matrices.
Each DNN layer contains one or more layer 2D  $N\times M$ weight matrices, $\mathbf{W}_{l}$, or pre-activation maps, $\mathbf{W}_{i,l}$, e.g., extracted from 2D Convolutional layers, where $N > M$.
See Appendix~\ref{sxn:appendix} for details.
(We may drop the $i$ and/or $i,l$ subscripts below.)
The best performing quality metrics depend on the norms and/or spectral properties of each weight matrix,
$\mathbf{W}$, and/or, equivalently, it's Empirical Correlation Matrix, $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$.
We consider the following metrics:
\begin{eqnarray}
& & \text{Frobenius Norm: $\Vert\mathbf{W}\Vert^{2}_{F}=\Vert\mathbf{X}\Vert_{F}=\sum\nolimits_{i=1}^{M} \lambda_{i}$ } \\
& & \text{Spectral Norm: $\Vert\mathbf{W}\Vert_{\infty}^{2}=\Vert\mathbf{X}\Vert_{\infty}=\lambda_{max}$ } \\
& & \text{Weighted Alpha: $\hat{\alpha}=\alpha\log\lambda_{max}$ } \\
& & \text{$\alpha$-Norm (or $\alpha$-Shatten Norm): $\Vert\mathbf{W}\Vert^{2\alpha}_{2\alpha}=\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}=\sum\nolimits_{i=1}^{M}\lambda_{i}^{\alpha}$. }
\end{eqnarray}
Here, $\lambda_{i}$ is the $i^{th}$ eigenvalue of the $\mathbf{X}$, and $\lambda_{max}$ is the maximum eigenvalue.
These eigenvalues are squares of the singular values $\sigma_{i}$ of $\mathbf{W}$, $\lambda_{i}=\sigma^{2}_{i}$.
%See Appendix~\ref{sxn:appendix} for details. 
All four metrics can be computed easily from DNN weight matrices.
The first two metrics are well-known in ML.
The last two metrics deserve special mention, as they depend on an empirical parameter $\alpha$ that is the PL exponent that arises in the recently-developed HT-SR Theory~\cite{MM18_TR, MM19_HTSR_ICML, MM20_SDM}.

\paragraph{Overview of Heavy-Tailed Self-Regularization.}

In the HT-SR Theory, one analyzes the eigenvalue spectrum, i.e., the Empirical Spectral Density (ESD), of the associated correlation matrices~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
From this, one characterizes the amount and form of correlation, and therefore implicit self-regularizartion, present in the DNN's weight matrices.
For each layer weight matrix $\mathbf{W}$, of size $N \times M$, construct the associated $M\times M$ (uncentered) correlation matrix $\mathbf{X}$. 
Dropping the $L$ and $l,i$ indices, one~has
$$
\mathbf{X} = \frac{1}{N}\mathbf{W}^{T}\mathbf{W}.
$$
If we compute the eigenvalue spectrum of $\mathbf{X}$, i.e., $\lambda_i$ such that
$  % $$
\mathbf{X}\mathbf{v}_{i}=\lambda_{i}\mathbf{v}_{i} , 
$  % $$
then the ESD of eigenvalues, $\rho(\lambda)$, is just a histogram of the eigenvalues, formally written as
%\begin{equation}
$\rho(\lambda)=\sum\nolimits_{i=1}^{M}\delta(\lambda-\lambda_{i})  .$
%\label{eqn:eigenval_hist}
%\end{equation}
Using HT-SR Theory, one characterizes the correlations in a weight matrix by examining its ESD, $\rho(\lambda)$.
It can be well-fit to a truncated PL distribution, given~as
\begin{equation}
\rho(\lambda)\sim\lambda^{-\alpha}  ,
\label{eqn:eigenval_pl}
\end{equation}
which is (at least) valid within a bounded range of eigenvalues $\lambda\in[\lambda^{min},\lambda^{max}]$.

The original work on HT-SR Theory considered a small number of NNs, including AlexNet and InceptionV3. 
It showed that for nearly every $\mathbf{W}$, the (bulk and tail) of the ESDs can be fit to a truncated PL, and that PL exponents $\alpha$ nearly all lie within the range $\alpha\in(1.5,5)$~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.
%
As for the mechanism responsible for these properties, statistical physics offers several possibilities~\cite{SornetteBook,nishimori01}, e.g., self organized criticality~\cite{SOC87,SOCat25yrs} or multiplicative noise in the stochastic optimization algorithms used to train these models~\cite{HodMah20A_TR,SorCon97}.
Our meta-analysis does not require knowledge of mechanisms; and it is not even clear that one mechanism is responsible for every case.
%
Crucially, HT-SR Theory predicts that smaller values of $\alpha$ should correspond to models with better correlation over multiple size scales and thus to better models.
Relatedly, previous work observed that smaller exponents $\alpha$ correspond to more implicit self-regularization and better generalization~\cite{MM18_TR,MM19_HTSR_ICML,MM20_SDM}.

\paragraph{DNN Empirical Quality Metrics.}

For norm-based metrics, we use the average of the log norm, to the appropriate power.
Informally, this amounts to assuming that the layer weight matrices are statistically independent, in which case we can estimate the model complexity $\mathcal{C}$, or test accuracy, with a standard Product Norm (which resembles a data dependent VC complexity),
\begin{equation}
\mathcal{C}\sim\Vert\mathbf{W}_{1}\Vert\times\Vert\mathbf{W}_{2}\Vert \times \cdots \times \Vert\mathbf{W}_{L}\Vert ,
\end{equation}
where $\Vert\cdot\Vert$ is a matrix norm.   
The log complexity,
\begin{equation}
\label{eqn:eqn:sum_log_norm}
\log\mathcal{C} \sim \log\Vert\mathbf{W}_{1}\Vert+\log\Vert\mathbf{W}_{2}\Vert + \cdots + \log\Vert\mathbf{W}_{L}\Vert = \sum\nolimits_l \log\Vert\mathbf{W}_{l}\Vert ,
\end{equation}
 takes the form of an average Log Norm.
For the \emph{Frobenius Norm metric} and \emph{Spectral Norm metric}, we can use Eqn.~(\ref{eqn:eqn:sum_log_norm}) directly (since, when taking $\log\Vert\mathbf{W}_{l}\Vert_{F}^{2}$, the $2$ comes down and out of the sum, and thus ignoring it only changes the metric by a constant factor).

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a much smaller and more limited set of models than we consider here.

Based on this, in this paper, we introduce and evaluate the \emph{$\alpha$-Shatten Norm metric},
\begin{equation}
\label{eqn:sum_log_alpha_norm_alpha}
\sum\nolimits_l \log \Vert\mathbf{X}_l\Vert_{\alpha_l}^{\alpha_l} 
=
\sum\nolimits_l \alpha_l \log \Vert\mathbf{X}_l\Vert_{\alpha_l} .
\end{equation}
For the $\alpha$-Shatten Norm metric, $\alpha_l$ varies from layer to layer, and so in Eqn.~(\ref{eqn:sum_log_alpha_norm_alpha}) it can not be taken out of the sum.
For small $\alpha$, the Weighted Alpha metric approximates the Log $\alpha$-Shatten norm, as can be shown with a statistical mechanics and random matrix theory derivation \cite{MM20_unpub_work}; and the Weighted Alpha and $\alpha$-Shatten norm metrics often behave like an improved, weighted average Log Spectral Norm.

One determines $\alpha$ for a given layer by fitting the ESD of that layer's weight matrix to a truncated PL, using the commonly accepted Maximum Likelihood method~\cite{CSN09_powerlaw,ABP14}.
This method works very well for exponents between $\alpha\in(2,4)$; and it is adequate, although imprecise, for smaller and especially larger $\alpha$~\cite{newman2005_zipf}. 
%
Operationally, $\alpha$ is determined by using the \emph{WeightWatcher} tool~\cite{weightwatcher_package} to fit the histogram of eigenvalues, $\rho(\lambda)$, to a truncated PL, 
\begin{equation}
\rho(\lambda)\sim\lambda^{\alpha},\;\;\lambda\in[\lambda_{min},\lambda_{max}] ,
\end{equation}
where $\lambda_{max}$ is the largest eigenvalue of $\mathbf{X}=\mathbf{W}^{T}\mathbf{W}$, and 
where $\lambda_{min}$ is selected automatically to yield the best (in the sense of minimizing the K-S distance) PL fit.
Each of these quantities is defined for a given layer $\mathbf{W}$ matrix.
See Figure~\ref{fig:ww} for an illustration.

\begin{figure}[t]
    \centering
    \includegraphics[width=15.0cm]{img/WeightWatcher_v2}
    \caption{Schematic of analyzing DNN layer weight matrices $\mathbf{W}$.  Norm-based metrics (not shown) and PL-based metrics (depending on fitting the histogram of eigenvalues to a truncated PL) can be used to compare models (or for other tasks such as improved model training).}
    \label{fig:ww}
\end{figure}

To avoid confusion, let us clarify the relationship between $\alpha$ and $\hat{\alpha}$.  
We fit the ESD of the correlation matrix $\mathbf{X}$ to a truncated PL, parameterized by 2 values: the PL exponent $\alpha$, and the maximum eigenvalue $\lambda_{max}$.
%%\footnote{Technically, we also need the minimum eigenvalue $\lambda_{min}$, but this detail does not affect our analysis.}
The PL exponent $\alpha$ measures the amount of correlation in a DNN layer weight matrix $\mathbf{W}$. 
It is valid for $\lambda\le\lambda_{max}$, and it is scale-invariant, i.e., it does not depend on the normalization of $\mathbf{W}$ or $\mathbf{X}$.
The $\lambda_{max}$ is a measure of the size, or scale, of $\mathbf{W}$.
Multiplying each $\alpha$ by the corresponding $\log\lambda_{max}$ weighs ``bigger'' layers more, and averaging this product leads to a balanced, Weighted Alpha metric $\hat{\alpha}$ for the entire~DNN.

>>>>

To be fully reproducible, we only examine publicly-available, pretrained models.
All of our computations were performed with the \texttt{WeightWatcher} tool (version 0.2.7)~\cite{weightwatcher_package}, and we provide all Jupyter and Google Colab notebooks used in an accompanying github repository~\cite{kdd20_sub_repo}, which includes more details and more results.

\paragraph{Additional Details on Layer Weight Matrices}

Recall that we can express the objective/optimization function for a typical DNN with $L$ layers, with activation functions $h_{l}(\cdot)$, and with $N\times M$ weight matrices $\mathbf{W}_{l}$ and biases $\mathbf{b}_{l}$ as Equation~(\ref{eqn:dnn_energy}).
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.
Typically, this model would be trained on some labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using Backprop, by minimizing the loss $\mathcal{L}$.
For simplicity, we do not indicate the structural details of the layers (e.g., Dense or not, Convolutions or not, Residual/Skip Connections, etc.). 
Each layer is defined by one or more layer 2D weight matrices $\mathbf{W}_{l}$, and/or the 2D feature maps $\mathbf{W}_{l,i}$ extracted from 2D Convolutional (Conv2D) layers.
A typical modern DNN may have anywhere between 5 and 5000 2D layer matrices.

For each Linear Layer, we get a  single $(N\times M)$ (real-valued) 2D weight matrix, denoted $\mathbf{W}_{l}$, for layer $l$.  
This includes Dense or Fully-Connected (FC) layers, as well as 1D Convolutional (Conv1D) layers, Attention matrices, etc.
We ignore the bias terms $\mathbf{b}_{l}$ in this analysis. 
Let the aspect ratio be $Q=\frac{N}{M}$, with $Q\ge 1$.
For the Conv2D layers, we have a 4-index Tensor, of the form $(N\times M \times c\times d)$, consisting
of $c\times d$ 2D feature maps of shape $(N\times M)$.    
We  extract $n_{l}=c\times d$ 2D weight matrices $\mathbf{W}_{l,i}$, one for each feature map $i=[1,\dots,n_{l}]$ for layer $l$.

\paragraph{SVD of Convolutional 2D Layers.}

There is some ambiguity in performing spectral analysis on Conv2D layers.  
Each layer is a 4-index tensor of dimension $(w,h,in,out)$, with an $(w\times h)$ filter (or kernel) and $(in, out)$
channels. When $w=h=k$, it gives $(k\times k)$ tensor slices, or pre-Activation Maps, $\mathbf{W}_{i,L}$ of dimension $(in\times out)$ each. 
%
We identify 3 different approaches for running SVD on a Conv2D layer:
\begin{enumerate}
\item run SVD on each pre-Activation Map $\mathbf{W}_{i,L}$, yielding $(k\times k)$ sets of $M$ singular values;
\item stack the maps into a single matrix of, say, dimension $((k\times k\times out)\times in)$, and run SVD to get $in$ singular values;
\item compute the 2D Fourier Transform (FFT) for each of the $(in, out)$ pairs, and run SVD on the Fourier coefficients~\cite{CNNSVD}, leading to $\sim(k\times in\times out)$ non-zero singular values.
\end{enumerate}
Each method has tradeoffs.  
Method (3) is mathematically sound, but computationally expensive. Method (2) is ambiguous.
For our analysis, because we need thousands of runs, we select method (1), which is the fastest (and is easiest to reproduce).

\paragraph{Normalization of Empirical Matrices.}

Normalization is an important, if underappreciated, practical issue.
Importantly, the normalization of weight matrices does \emph{not} affect the PL fits because $\alpha$ is scale-invariant.
Norm-based metrics, however, do depend strongly on the scale of the weight matrix---that is the point.
To apply RMT, we usually define $\mathbf{X}$ with a $1/N$ normalization, assuming variance of $\sigma^{2}=1.0$.
Pretrained DNNs are typically initialized with random weight matrices $\mathbf{W}_{0}$, with $\sigma^{2}\sim 1/\sqrt{N}$, or some variant, e.g., the Glorot/Xavier normalization~\cite{GloBen10}, or a $\sqrt{2/Nk^2}$ normalization for Convolutional 2D Layers. With this implicit scale, we do \emph{not} ``renormalize'' the empirical weight matrices, i.e., we use them as-is.
The only exception is that \emph{we do rescale} the Conv2D pre-activation maps $\mathbf{W}_{i,L}$ by $k/\sqrt{2}$ so that they are on the same scale as the Linear / Fully Connected (FC) layers.

\paragraph{Special consideration for NLP models.}

NLP models, and other models with large initial embeddings, require special care because the embedding layers frequently lack the implicit $1/\sqrt{N}$ normalization present in other layers.
For example, in GPT, for most layers, the maximum eigenvalue $\lambda_{max}\sim\mathcal{O}(10-100)$, but in the first embedding layer, the maximum eigenvalue is of order $N$ (the number of words in the embedding), or $\lambda_{max}\sim\mathcal{O}(10^{5})$.  
For GPT and GPT2, we treat all layers as-is (although one may want to normalize the first 2 layers $\mathbf{X}$ by $1/N$, or to treat them as outliers).

### 2020-10-01 submission/NatCom_nlp_models.tex
pass to address most of the presubmission comments for nat comm

Within the past few years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the information correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
More generally, compared to CV models,
modern NLP models have larger weight matrices and display different spectral properties.

@@

\paragraph{What do large values of $\alpha$ mean?}

Many NLP models, such as early variants of GPT and BERT, have weight matrices with unusually large PL exponents (e.g., $\alpha\gg 6$).
This indicates these matrices may be \emph{under}-correlated (i.e., over-parameterized, relative to the amount of data).
In this regime, the truncated PL fit itself may not be very reliable because the Maximum Likelihood estimator it uses is unreliable in this range.
The specific $\alpha$ values returned by the truncated PL fits are less reliable, but having large versus small $\alpha$ is reliable.
If the ESD is visually examined, one can usually describe these $\mathbf{W}$ as in the \emph{Bulk-Decay} or \emph{Bulk-plus-Spikes} phase from HT-ST Theory~\cite{MM18_TR,MM19_HTSR_ICML}.
Previous work~\cite{MM18_TR,MM19_HTSR_ICML} has conjectured that very well-trained DNNs would not have many \emph{outlier} $\alpha\gg 6$.
Consistent with this, more recent improved versions of GPT (shown below) and BERT (not shown) confirm this.

@@

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
\hline
 Series  & \#   & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\hat{\alpha}$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT2-small & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
\hline
GPT2-medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2-large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2-xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the average Log Norm and Weighted Alpha metrics for pretrained OpenAI GPT and GPT2 models. 
Column \# refers to number of layers treated.  
Averages do not include the first embedding layer(s) because they are not (implicitly) normalized.  
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrices.  
Self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight matrices.
}
\label{table:nlp}
\end{table}

@@

See Table \ref{table:nlp} for a summary of results for the four quality metrics for the OpenAI GPT and GPT2 pretrained models.
Comparing trends between GPT and GPT2-small, 
observe that all four metrics increase when going from GPT to GPT2-small, i.e., they are smaller for the higher-quality model (higher quality since GPT was trained to better data), when the number of layers is held fixed.
In the poorly-trained GPT model, the norm metrics all exhibit \emph{Scale Collapse}, compared to well-trained GPT2-small model.

Comparing trends between GPT2-medium to GPT2-large to GPT2-xl,
observe that (with one minor exception involving the log Frobenius norm metric) all four metrics decrease as one goes from medium to large to xl, indicating that the larger models indeed look better than the smaller models.
For these well-trained models, the norm metrics now behave as expected, i.e., there is no \emph{Scale Collapse} and norms are decreasing with increasing~accuracy.

@@

We next describe the Spectral norm in GPT versus GPT2-small.
In Figure~\ref{fig:GPT-snorm-hist}, the poorly-trained GPT model has a smaller mean/median Spectral norm as well as, spuriously, many much smaller Spectral norms, compared to the well-trained GPT2-small.
This violates the conventional wisdom that smaller Spectral norms are better.
Because there are so many anonymously small Spectral norms, the GPT model may be exhibiting a kind of \emph{Scale Collapse}, like that observed in the distilled CV models (Figure~\ref{fig:resnet204D5L}).

This demonstrates that, while the Spectral (or Frobenius) norm may correlate well with predicted test error, at least among reasonably well-trained models, it is \emph{not} a good indicator of the overall model quality in general.
It can mispredict well-trained versus poorly-trained questions in ways not seen with PL-based metrics.
Using it as an empirical quality metric may give spurious results when applied to poorly-trained or otherwise deficient~models.

\begin{figure}[h]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    \qquad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        \includegraphics[width=5.5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Histogram of PL exponents ($\alpha$) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.}

@@

Figure~\ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms.
Upon examination, e.g., from Figure \ref{fig:gpt-snorm-layer} (below), these correspond to the first embedding layer(s).
These layers have a different effective normalization, and therefore a different scale.
See Appendix~\ref{sxn:appendix} for details.
Here, we do not include them in our computed average metrics in Table~\ref{table:nlp}, and we do not include them in the histogram plot in Figure~\ref{fig:GPT-snorm-hist}.

@@

\begin{figure}[htb]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \quad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        \includegraphics[width=5.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{PL exponents ($\alpha$) (in (a)) and Log Spectral Norms ($\log\Vert\mathbf{W}\Vert_{\infty}$) (in (b)) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.  (Note that the quantities being shown on each Y axis are different.)
             In the text, this is interpreted in terms of \emph{Correlation Flow} and \emph{Scale~Collapse}.
            }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    \quad
    \subfigure[Log Alpha Norm]{
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Histogram of PL exponents ($\alpha$)
              and Log Alpha Norm ($\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$) for weight matrices from models of different sizes in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anomalously large values.)
             }
    \label{fig:gpt2-histograms}
\end{figure}

>>>>

Within the past few years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the information correlations in the data (relative to their size) and thus are substantially under-trained.
More generally, compared to CV models,
modern NLP models have larger weight matrices and display different spectral properties.

@@

Many NLP models, such as early variants of GPT and BERT, have weight matrices with unusually large PL exponents (e.g., $\alpha\gg 6$).
This indicates these matrices may be under-correlated (i.e., over-parameterized, relative to the amount of data).
In this regime, the truncated PL fit itself may not be very reliable because the Maximum Likelihood estimator it uses is unreliable in this range.
The specific $\alpha$ values returned by the truncated PL fits are less reliable, but having large versus small $\alpha$ is reliable.
If the ESD is visually examined, one can usually describe these $\mathbf{W}$ as in the Bulk-Decay or Bulk-plus-Spikes phase from HT-ST Theory~\cite{MM18_TR,MM19_HTSR_ICML}.
Previous work~\cite{MM18_TR,MM19_HTSR_ICML} has conjectured that very well-trained DNNs would not have many outlier $\alpha\gg 6$.
Consistent with this, more recent improved versions of GPT (shown below) and BERT (not shown) confirm~this.

@@

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
\hline
 Series  & \#   & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\hat{\alpha}$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT2-small & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
\hline
GPT2-medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2-large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2-xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the average Log Norm and Weighted Alpha metrics for pretrained OpenAI GPT and GPT2 models. 
Column \# refers to number of layers treated.  
Averages do not include the first embedding layer(s) because they are not (implicitly) normalized.  
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrices.  
Self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight~matrices.
}
\label{table:nlp}
\end{table}

@@

See Table \ref{table:nlp} for a summary of results for the four quality metrics for the OpenAI GPT and GPT2 pretrained models.
Comparing trends between GPT and GPT2-small, 
observe that all four metrics increase when going from GPT to GPT2-small, i.e., they are smaller for the higher-quality model (higher quality since GPT was trained to better data), when the number of layers is held fixed.
In the poorly-trained GPT model, the norm metrics all exhibit Scale Collapse, compared to well-trained GPT2-small model.

Comparing trends between GPT2-medium to GPT2-large to GPT2-xl,
observe that (with one minor exception involving the log Frobenius norm metric) all four metrics decrease as one goes from medium to large to xl, indicating that the larger models indeed look better than the smaller models.
For these well-trained models, the norm metrics now behave as expected, i.e., there is no Scale Collapse and norms are decreasing with increasing~accuracy.

@@

We next describe the Spectral norm in GPT versus GPT2-small.
In Figure~\ref{fig:GPT-snorm-hist}, the poorly-trained GPT model has a smaller mean/median Spectral norm as well as, spuriously, many much smaller Spectral norms, compared to the well-trained GPT2-small.
This violates the conventional wisdom that smaller Spectral norms are better.
Because there are so many anonymously small Spectral norms, the GPT model may be exhibiting a kind of Scale Collapse, like that observed in the distilled CV models (Figure~\ref{fig:resnet204D5L}).

This demonstrates that, while the Spectral (or Frobenius) norm may correlate well with predicted test error, at least among reasonably well-trained models, it is not a good indicator of the overall model quality in general.
Using it as an empirical quality metric may give spurious results when applied to poorly-trained or otherwise deficient~models.

\begin{figure}[h]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT-alpha-hist.png}
        \label{fig:GPT-alpha-hist}
       }
    \qquad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        \includegraphics[width=5.5cm]{img/GPT-snorm-hist.png}
        \label{fig:GPT-snorm-hist}
       }
   \caption{Histogram of PL exponents 
           %($\alpha$) 
           and Log Spectral Norms 
           %($\log\Vert\mathbf{W}\Vert_{\infty}$) 
           for weight matrices from the OpenAI GPT and GPT2-small pretrained models.}

@@

Figure~\ref{fig:GPT-snorm-hist} also shows some unusually large Spectral Norms.
Upon examination, e.g., from Figure \ref{fig:gpt-snorm-layer} (below), these correspond to the first embedding layer(s).
These layers have a different effective normalization, and therefore a different scale.
See the Supplementary Information
%~\ref{sxn:appendix} 
for details.
Here, we do not include them in our computed average metrics in Table~\ref{table:nlp}, and we do not include them in the histogram plot in Figure~\ref{fig:GPT-snorm-hist}.

@@

\begin{figure}[htb]
    \centering
    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT-alpha-depth.png}
        \label{fig:gpt-alpha-layer}
    }
    \quad
    \subfigure[Log Spectral Norm ($\log\Vert\mathbf{W}\Vert_{\infty}$)]{
        \includegraphics[width=5.5cm]{img/GPT-snorm-depth.png}
        \label{fig:gpt-snorm-layer}
    }
    \caption{PL exponents 
             %($\alpha$) 
             (in (a)) and Log Spectral Norms 
             %($\log\Vert\mathbf{W}\Vert_{\infty}$) 
             (in (b)) for weight matrices from the OpenAI GPT and GPT2-small pretrained models.  (Note that the quantities being shown on each Y axis are different.)
             In the text, this is interpreted in terms of Correlation Flow and Scale~Collapse.
            }
    \label{fig:gpt-alpha-layers}
\end{figure}

@@

    \subfigure[PL exponent ($\alpha$)]{
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    \quad
    \subfigure[Log Alpha Norm]{
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Histogram of PL exponents 
             %($\alpha$)
             and Log Alpha Norm 
             %($\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$) 
             for weight matrices from models of different sizes in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anomalously large values.)
             }
    \label{fig:gpt2-histograms}
\end{figure}
